---
title: "Single Cell Data Set - Lake 2018"
output: 
  md_document
---

# Load Data

### Load single cell dataset

```{R, load data,message=FALSE, warning=FALSE}
library(tidyverse)
```

# Compute sum of UMIs

Cerebellum ce ll types

```{r,eval=FALSE}
cerebellum <- read_tsv("../../Data/Lake2018/GSE97930_CerebellarHem_snDrop-seq_UMI_Count_Matrix_08-01-2017.txt")
annot <- tibble(cell=colnames(cerebellum)[-1],cell_type=gsub("_cbm[0-9]_[ATGCN]+$","",colnames(cerebellum)[-1]))

cerebellum <- gather(cerebellum,cell,UMI,Gran_cbm1_TTAATCAGTCGC:Mic_cbm9_ATGGTCGGAGAC,-Gene)
cerebellum <- inner_join(cerebellum,annot)

cerebellum <- cerebellum %>% group_by(Gene,cell_type) %>% summarise(sumUMI=sum(UMI))
write_tsv(cerebellum,path = "../Data/Lake2018/GSE97930_CerebellarHem_snDrop-seq_UMI_Count_Matrix_08-01-2017_sumUMI.txt")
```

Frontal cortex cell types 

```{r,eval=FALSE}
fc <- read_tsv("../../Data/Lake2018/GSE97930_FrontalCortex_snDrop-seq_UMI_Count_Matrix_08-01-2017.txt")

fc <- gather(fc,cell,UMI,Ex1_fcx8_GGACGCCTTTAA:Mic_fcx13_GGGAGTAACTTG,-Gene)
cell_type1M <- gsub("_fcx[0-9]{1,2}_[ATGCN]+$","",fc$cell[1:100000000])
cell_type2M <- gsub("_fcx[0-9]{1,2}_[ATGCN]+$","",fc$cell[100000001:200000000])
cell_type3M <- gsub("_fcx[0-9]{1,2}_[ATGCN]+$","",fc$cell[200000001:353993295])
cell_types <- c(cell_type1M,cell_type2M,cell_type3M)
fc$cell_type <- cell_types

fc <- fc %>% group_by(Gene,cell_type) %>% summarise(sumUMI=sum(UMI))
write_tsv(fc,path = "../../Data/Lake2018/GSE97930_FrontalCortex_snDrop-seq_UMI_Count_Matrix_08-01-2017_sumUMI.txt")
```

Visaul cortex cell types 

```{r,eval=FALSE}
vc <- read_tsv("../../Data/Lake2018/GSE97930_VisualCortex_snDrop-seq_UMI_Count_Matrix_08-01-2017.txt")

vc <- gather(vc,cell,UMI,Ex1_occ17_AAGTGAGTGACC:Mic_occ24_ATTAATTGAAGA,-Gene)
cell_type1M <- gsub("_occ[0-9]{1,2}_[ATGCN]+$","",vc$cell[1:100000000])
cell_type2M <- gsub("_occ[0-9]{1,2}_[ATGCN]+$","",vc$cell[100000001:200000000])
cell_type3M <- gsub("_occ[0-9]{1,2}_[ATGCN]+$","",vc$cell[200000001:300000000])
cell_type4M <- gsub("_occ[0-9]{1,2}_[ATGCN]+$","",vc$cell[300000001:400000000])
cell_type5M <- gsub("_occ[0-9]{1,2}_[ATGCN]+$","",vc$cell[400000001:500000000])
cell_type6M <- gsub("_occ[0-9]{1,2}_[ATGCN]+$","",vc$cell[500000001:600000000])
cell_type7M <- gsub("_occ[0-9]{1,2}_[ATGCN]+$","",vc$cell[600000001:700000000])
cell_type8M <- gsub("_occ[0-9]{1,2}_[ATGCN]+$","",vc$cell[700000001:804566088])


cell_types <- c(cell_type1M,cell_type2M,cell_type3M,cell_type4M,cell_type5M,cell_type6M,cell_type7M,cell_type8M)
vc$cell_type <- cell_types

vc <- vc %>% group_by(Gene,cell_type) %>% summarise(sumUMI=sum(UMI))
write_tsv(vc,path = "../../Data/Lake2018/GSE97930_VisualCortex_snDrop-seq_UMI_Count_Matrix_08-01-2017_sumUMI.txt")
```

# Load computed UMI for each tissue

```{r}
cerebellum <- read_tsv("../../Data/Lake2018/GSE97930_CerebellarHem_snDrop-seq_UMI_Count_Matrix_08-01-2017_sumUMI.txt")
cerebellum <- mutate(cerebellum,tissue="cerebellum")
```

```{r}
fc <- read_tsv("../../Data/Lake2018/GSE97930_FrontalCortex_snDrop-seq_UMI_Count_Matrix_08-01-2017_sumUMI.txt")
fc <- mutate(fc,tissue="Frontal Cortex")
```

```{r}
vc <- read_tsv("../../Data/Lake2018/GSE97930_VisualCortex_snDrop-seq_UMI_Count_Matrix_08-01-2017_sumUMI.txt")
vc <- mutate(vc,tissue="Visual Cortex")
```

Join the cell types from the three tissues

```{r}
exp <- rbind(cerebellum,fc,vc)
```

Get sum of UMI for each gene and each cell type with the same name

```{r}
exp_lvl5 <- exp %>% group_by(Gene,cell_type) %>% summarise(sumUMI=sum(sumUMI)) %>% ungroup()
```

Tidy the data

```{r}
exp_lvl5 <- exp_lvl5 %>% spread(cell_type,sumUMI)
exp_lvl5[is.na(exp_lvl5)] <- 0
exp_lvl5 <- gather(exp_lvl5,cell_type,sumUMI,Ast:Purk2)
```

### Load gene coordinates

Load gene coordinates and extend upstream and downstream by 100kb.

File downloaded from MAGMA website (https://ctg.cncr.nl/software/magma).

Filtered to remove extended MHC (chr6, 25Mb to 34Mb).

```{R, load coordinates,message=F}
gene_coordinates <- read.table("../../Data/NCBI/NCBI37.3.gene.loc.extendedMHCexcluded",header=F,stringsAsFactors = F) %>% 
  mutate(start=ifelse(V3-100000<0,0,V3-100000),end=V4+100000,V1=as.character(V1)) %>% 
  select(2,start,end,1) %>% 
  as.tibble() %>% 
  rename(chr="V2", ENTREZ="V1") %>% 
  mutate(chr=paste0("chr",chr))
```

### Write dictonary for cell type names

```{R, write dictionnary,message=F}
exp_lvl5 <- exp_lvl5 %>% ungroup() %>% rename(Lvl5=cell_type,Expr_sum_mean=sumUMI)
dic_lvl5 <- select(exp_lvl5,Lvl5) %>% unique() %>% mutate(makenames=make.names(Lvl5))
write_tsv(dic_lvl5,"../../Data/Lake2018/dictionary_cell_type_names.txt")
```

Remove not expressed genes

```{r}
genes_to_remove <- exp_lvl5 %>% group_by(Gene) %>% summarise(sum=sum(Expr_sum_mean)) %>% filter(sum==0)
exp_lvl5 <- filter(exp_lvl5,!Gene%in%genes_to_remove$Gene)
```

# QC

Remove not expressed genes

```{r}
genes_to_remove <- exp_lvl5 %>% 
  group_by(Gene) %>% 
  summarise(sum=sum(Expr_sum_mean)) %>% 
  filter(sum==0)
exp_lvl5 <- filter(exp_lvl5,!Gene%in%genes_to_remove$Gene)
```

### Scale to 1 million molecules

Each cell type is scaled to the same total number of molecules. 

```{R, rm lowQ 34,message=F}
exp_lvl5 <- exp_lvl5 %>% 
  group_by(Lvl5) %>% 
  mutate(Expr_sum_mean=Expr_sum_mean*1e6/sum(Expr_sum_mean))
```

# Specificity Calculation

The specifitiy is defined as the proportion of total expression performed by the cell type of interest (x/sum(x)).

### Lvl5

```{R, specificity 5,message=F}
exp_lvl5 <- exp_lvl5 %>% group_by(Gene) %>% 
  mutate(specificity=Expr_sum_mean/sum(Expr_sum_mean))
```

### Keep only ENTREZ genes protein coding and MAGMA tested genes

```{r}
entrez2symbol <- AnnotationDbi::toTable(org.Hs.eg.db::org.Hs.egSYMBOL2EG) %>% rename(Gene="symbol",ENTREZ="gene_id")
exp_lvl5 <- inner_join(exp_lvl5,entrez2symbol,by="Gene") 
exp_lvl5 <- inner_join(exp_lvl5,gene_coordinates,by="ENTREZ") 
```

### Get number of genes

Get number of genes that represent 10% of the dataset

```{r}
n_genes <- length(unique(exp_lvl5$ENTREZ))
n_genes_to_keep <- (n_genes * 0.1) %>% round()
```

# Save expression profile for other processing

```{r}
save(exp_lvl5,file = "expression.ready.Rdata")
```

### Functions

#### Get MAGMA input top10%

```{R, Functions 6,message=F}
magma_top10 <- function(d,Cell_type){
  d_spe <- d %>% group_by_(Cell_type) %>% top_n(.,n_genes_to_keep,specificity) 
  d_spe %>% do(write_group_magma(.,Cell_type))
}
```

```{R, write_function1}
write_group_magma  = function(df,Cell_type) {
  df <- select(df,Lvl5,ENTREZ)
  df_name <- make.names(unique(df[1]))
  colnames(df)[2] <- df_name  
  dir.create(paste0("MAGMA/"), showWarnings = FALSE)
  select(df,2) %>% t() %>% as.data.frame() %>% rownames_to_column("Cat") %>%
  write_tsv("MAGMA/top10.txt",append=T)
return(df)
}
```

#### Get LDSC input top 10%

```{R, write_function}
write_group  = function(df,Cell_type) {
  df <- select(df,Lvl5,chr,start,end,ENTREZ)
  dir.create(paste0("LDSC/Bed"), showWarnings = FALSE,recursive = TRUE)
  write_tsv(df[-1],paste0("LDSC/Bed/",make.names(unique(df[1])),".bed"),col_names = F)
return(df)
}
```

```{R, Functions 23,message=F}
ldsc_bedfile <- function(d,Cell_type){
  d_spe <- d %>% group_by_(Cell_type) %>% top_n(.,n_genes_to_keep,specificity) 
  d_spe %>% do(write_group(.,Cell_type))
}
```

### Write MAGMA/LDSC input files 

Filter out genes with expression below 1 TPM.

```{R, magma input21211223123,message=F, results='hide'}
exp_lvl5 %>% filter(Expr_sum_mean>1) %>% magma_top10("Lvl5")
```

```{R, ldsc22223123,message=F, results='hide'}
exp_lvl5 %>% filter(Expr_sum_mean>1) %>% ldsc_bedfile("Lvl5")
```